Discrete Quantum Theories and Computing

### 2018-03-04 dissertation.tex
dissertation.lyx "2.1 Geometrical Structure of States" Updating...

There are many things that are assumed in CQT, such as the absence
of zero norm states for non-zero vectors, and the decomposition of
complex amplitudes into a pair of ordinary real numbers. One also
typically assumes the existence of a $D$-dimensional Hilbert space
with an orthonormal basis, allowing us to write \textit{pure} states
in general as Hilbert space vectors with an Hermitian inner product:
\begin{eqnarray}
\ket{\Psi} & = & \sum_{i=0}^{D-1}\alpha_{i}\ket{i}\,.
\end{eqnarray}
Here $\alpha_{i}\in\mathComplex$ are complex probability amplitudes,
$\VVec{\alpha}\in\mathComplex^{D}$, and the $\{\ket{i}\}$ is an
orthonormal basis of states obeying 
\begin{equation}
\braket{i}{k}=\delta_{ik}\,.
\end{equation}

@@

A state in a two-dimensional Hilbert space, also called a qubit, already
provides access to a wealth of geometric information and context.
We write the single qubit state as 
\begin{align}
\ket{\psi_{1}} & =\alpha_{0}\ket{0}+\alpha_{1}\ket{1} & \alpha_{0},\alpha_{1} & \in\mathbb{C}\,.\label{A1.eq}
\end{align}

A convenience for computing probability and relative state properties
is the normalization condition 
\begin{equation}
\|\psi_{1}\|^{2}=|\alpha_{0}|^{2}+|\alpha_{1}|^{2}=\alpha_{0}^{*}\alpha_{0}+\alpha_{1}^{*}\alpha_{1}=1\ ,\label{A2.eq}
\end{equation}
which identifies $\alpha_{0}$ and $\alpha_{1}$ as (complex) probability
amplitudes and implies the conservation of probability in the closed
world spanned by $\{\ket{0},\ket{1}\}$. Note that we distinguish
for future use the \textit{norm\/} $\|\cdot\|$ of a vector from
the \textit{modulus\/} $|\cdot|$ of a complex number. Continuing,
we see that if we want only the irreducible state descriptions, we
must supplement the process of computing \Eq{A2.eq} by finding a
way to remove the distinction between states that differ only by an
overall phase transformation $\rme^{\rmi\frac{\theta}{2}}$, that
is, 
\begin{equation}
\left(\alpha_{0},\alpha_{1}\right)\sim\left(\rme^{\rmi\frac{\theta}{2}}\alpha_{0},\rme^{\rmi\frac{\theta}{2}}\alpha_{1}\right)\,.\label{A3.eq}
\end{equation}
This can be accomplished by the Hopf fibration \cite{Hatcher2001,MosseriDandoloff2001,Bengtsson2007,wiki:HopfFibration},
which we can write down as follows: let 
\begin{align}
\alpha_{0} & =x_{0}+\rmi y_{0}\,, & \alpha_{1} & =x_{1}+\rmi y_{1}\,.\label{A4.eq}
\end{align}
Then \Eq{A2.eq} becomes the condition that the four real variables
describing a qubit denote a point on the three-sphere $\Sphere{3}$
(a 3-manifold) embedded in $\R^{4}$: 
\begin{equation}
{x_{0}}^{2}+{y_{0}}^{2}+{x_{1}}^{2}+{y_{1}}^{2}=1\ .\label{A5.eq}
\end{equation}

We can reduce 3 real degrees of freedom in \Eq{A5.eq} to 2 real
degrees of freedom by effectively removing $\rme^{\rmi\frac{\theta}{2}}$
(``fibering out by the circle $\Sphere{1}$''). The standard form
of this maps (``the Hopf fibration'') is 
\begin{eqnarray}
X & = & 2\,\mathrm{Re}\ \alpha_{0}\alpha_{1}^{*}=2x_{0}x_{1}+2y_{0}y_{1}\,,\nonumber \\
Y & = & 2\,\mathrm{Im}\ \alpha_{0}\alpha_{1}^{*}=2x_{1}y_{0}-2x_{0}y_{1}\,,\label{eq:HopfFibration}\\
Z & = & |\alpha_{0}|^{2}-|\alpha_{1}|^{2}={x_{0}}^{2}+{y_{0}}^{2}-{x_{1}}^{2}-{y_{1}}^{2}\,.\nonumber 
\end{eqnarray}
By denoting the the three-dimensional vector~$\left(X,Y,Z\right)$
as $\Hat{a}$, \Eq{A5.eq} implies these transformed coordinates
obeying 
\begin{equation}
\|\Hat{a}\|^{2}=X^{2}+Y^{2}+Z^{2}=\left(|\alpha_{0}|^{2}+|\alpha_{1}|^{2}\right)^{2}=1\label{A7.eq}
\end{equation}
and therefore have only two remaining degrees of freedom describing
all possible distinct one-qubit quantum states. In Fig.~\ref{hopfFibers.fig},
we illustrate schematically the family of circles \textit{each one
of which is collapsed to a point\/} $\left(\phi,\psi\right)$ on
the surface $X^{2}+Y^{2}+Z^{2}=1$ by the Hopf map.

\begin{figure}[tbh]
\hfill{}\subfloat[]{\centering{}\includegraphics[width=0.48\columnwidth]{hopfS2}}\hfill{}\subfloat[]{\begin{centering}
\includegraphics[width=0.48\columnwidth]{hopfFibers}
\par\end{centering}
}\hfill{} \caption{(a) The sphere represented by \Eq{A7.eq}, which is the irreducible
space of one-qubit states, along with a representative set of points
on the sphere. (b) Representation of the Hopf fibration, \ref{eq:HopfFibration},
as a family of circles (the paths of $\rme^{\rmi\frac{\theta}{2}}$)
on the three-sphere $\Sphere{3}$. Although $\Sphere{3}$ cannot be
directly embedded in $\R^{3}$, $\Sphere{3}$ can be regarded as attaching
two three-dimensional ball on two sides of two-sphere $\Sphere{2}$.
In this way, each circle in $\Sphere{3}$ corresponding to a single
point on the sphere in (a) can be represented as a circle in the three-dimensional
ball. Moreover, points in (a) are color coded corresponding to circles
in (b), e.g., one pole contains the red elliptical circle that would
become an infinite-radius circle by a slightly different way to represent
$\Sphere{3}$ in $\R^{3}$, and the opposite pole corresponds to the
large perfectly round red circle at the equator.}
\label{hopfFibers.fig} 
\end{figure}

The resulting manifold is the two-sphere $\Sphere{2}$ (a 2-manifold)
embedded in $\R^{3}$. If we choose one of many possible coordinate
systems describing $\Sphere{3}$ via \Eq{A5.eq} such as 
\begin{equation}
\left(x_{0},y_{0},x_{1},y_{1}\right)=\left(\cos\frac{\theta+\phi}{2}\cos\frac{\psi}{2},\,\sin\frac{\theta+\phi}{2}\cos\frac{\psi}{2},\,\cos\frac{\theta-\phi}{2}\sin\frac{\psi}{2},\,\sin\frac{\theta-\phi}{2}\sin\frac{\psi}{2}\right)\,,\label{A8a.eq}
\end{equation}
where %% $0\leq \theta \leq 2 \pi$,  $-\pi \leq \phi \leq  \pi$, 
$0\leq\psi\leq\pi$, with $0\leq\frac{\theta+\phi}{2}<2\pi$ and $0\leq\frac{\theta-\phi}{2}<2\pi$,
we see that 
\begin{equation}
\left(X,Y,Z\right)=\left(\cos\phi\sin\psi,\,\sin\phi\sin\psi,\,\cos\psi\right)\,.\label{A8b.eq}
\end{equation}
Thus the one-qubit state is independent of $\theta$, and we can choose
$\theta=\phi$ without loss of generality, reducing the form of the
unique one-qubit states to 
\begin{equation}
\ket{\psi_{1}}=\rme^{\rmi\phi}\cos\frac{\psi}{2}\ket{0}+\sin\frac{\psi}{2}\ket{1}\,,\label{A9.eq}
\end{equation}
and an irreducible state can be represented as a point on a sphere
called the Bloch sphere, as shown in Fig.~\ref{blochSphere.fig}(a).

\begin{figure}[thb]
\begin{centering}
\hfill{}\subfloat[]{\centering{}\includegraphics[width=0.47\columnwidth]{blochSphere}}\hfill{}\subfloat[]{\begin{centering}
\includegraphics[width=0.47\columnwidth]{blochSphereArc}
\par\end{centering}
}\hfill{}
\par\end{centering}
\caption{(a) The conventional Bloch sphere with a unique state represented
by the point at the red sphere. (b) The geodesic shortest-distance
arc connecting two one-qubit quantum states.}
\label{blochSphere.fig} 
\end{figure}

@@

For one qubit, we could easily solve the problem of reducing the full
unit-norm space to its irreducible components $\Vec{X}=(X,Y,Z)$ characterizing
the Bloch sphere. We have just argued that essentially the same process
is possible for $n$-qubits: in the abstract argument, we simply identify
the family of coefficients $\{\alpha_{i}\}$ as being the same if
they differ only by an overall phase $\rme^{\rmi\frac{\theta}{2}}$.
However, in practice this is not a construction that is easy to realize
in a practical computation. We now outline an explicit algorithm for
accomplishing the reduction to the irreducible $n$-qubit state space
$\CP{D-1}$; this construction will turn out to be useful for the
validation of our discrete results to follow below.

@@

We can now use the complex generalization of the classical Veronese
coordinate system for projective geometry to remove the overall phase
ambiguity $\rme^{\rmi\frac{\theta}{2}}$ from the $n$-qubit states.
If we take a particular weighting of the elements of the density matrix
$\rho$, we can construct a \textit{unit vector\/} of real dimension
$D^{2}$ with the form: 
\begin{equation}
\Vec{X}=(|\alpha_{i}|^{2},\ldots,\sqrt{2}\,\mathrm{Re}\ \alpha_{i}\alpha_{j}^{*},\ldots,\sqrt{2}\,\mathrm{Im}\ \alpha_{i}\alpha_{j}^{*},\ldots)\ ,\label{nqCoords.eq}
\end{equation}
where 
\begin{eqnarray}
\Vec{X}\cdot\Vec{X}=\left(\sum_{i=0}^{D-1}\left|\alpha_{i}\right|^{2}\right)^{2}=1\ .
\end{eqnarray}
This construction gives an explicit embedding of the $(D-1)$-dimensional
complex, or $(2D-2)=(2^{n+1}-2)$-dimensional real, object in a real
space of dimension $D^{2}=2^{2n}$. However, this is somewhat subtle
because the vector is of unit length, so technically the embedding
space is a sphere of dimension $D^{2}-1=2^{2n}-1$ embedded in $\R^{D^{2}}$;
the one-qubit irreducible states could be represented in a 4D embedding,
but the magnitude of every coordinate would be one; furthermore, the
object embedded in the resulting $\Sphere{3}$ is indeed $\Sphere{2}$
because we can fix one complex coordinate to be unity, and let one
vary, giving a total of two irreducible dimensions. In fact one must
choose \textit{two\/} coordinate patches, one covering one pole of
$\Sphere{2}$ with coordinates 
\begin{eqnarray}
\begin{array}{ccc}
\alpha_{0} & = & 1+0i\\
\alpha_{1} & = & x_{1}+iy_{1}
\end{array}
\end{eqnarray}
and the other patch covering the other pole of $\Sphere{2}$ with
coordinates 
\begin{eqnarray}
\begin{array}{ccc}
\alpha_{0} & = & x_{0}+iy_{0}\\
\alpha_{1} & = & 1+0i\ .
\end{array}
\end{eqnarray}

>>>>

There are many things that are assumed in CQT, such as the absence
of zero norm states for non-zero vectors, and the decomposition of
complex amplitudes into a pair of ordinary real numbers. One also
typically assumes the existence of a $D$-dimensional Hilbert space
with an orthonormal basis, allowing us to write \textit{pure} states
in general as Hilbert space vectors with an Hermitian inner product:
\begin{eqnarray}
\ket{\Psi} & = & \sum_{i=0}^{D-1}\alpha_{i}\ket{i}\,.
\end{eqnarray}
Here $\alpha_{i}\in\mathComplex$ are complex probability amplitudes,
$\VVec{\alpha}\in\mathComplex^{D}$, and the $\{\ket{i}\}$ is an
orthonormal basis of states obeying $\braket{i}{k}=\delta_{ik}$.

@@

A state in a two-dimensional Hilbert space, known as a qubit, already
provides access to a wealth of geometric information and context.
When we write the single qubit state as $\ket{\psi_{1}}=\alpha_{0}\ket{0}+\alpha_{1}\ket{1}$,
a convenience for computing probability and relative state properties
is the normalization condition 
\begin{equation}
\left\Vert \psi_{1}\right\Vert ^{2}=\left|\alpha_{0}\right|^{2}+\left|\alpha_{1}\right|^{2}=\alpha_{0}^{*}\alpha_{0}+\alpha_{1}^{*}\alpha_{1}=1\,,\label{A2.eq}
\end{equation}
which identifies $\alpha_{0}$ and $\alpha_{1}\in\mathbb{C}$ as probability
amplitudes and implies the conservation of probability in the closed
world spanned by $\left\{ \ket{0},\ket{1}\right\} $. Note that we
distinguish for future use the \textit{norm}~$\left\Vert \cdot\right\Vert $
of a vector from the \textit{modulus}~$\left|\cdot\right|$ of a
complex number. Continuing, we see that if we want only the irreducible
state descriptions, we must supplement the process of computing \Eq{A2.eq}
by finding a way to remove the distinction between states that differ
only by an overall phase transformation $\rme^{\rmi\theta}$, that
is, $\left(\alpha_{0},\alpha_{1}\right)\sim\left(\rme^{\rmi\theta}\alpha_{0},\rme^{\rmi\theta}\alpha_{1}\right)$.
This can be accomplished by the Hopf fibration \cite{Hatcher2001,MosseriDandoloff2001,Bengtsson2007,wiki:HopfFibration},
which can be written down as follows: let $\alpha_{0}=x_{0}+\rmi y_{0}$
and $\alpha_{1}=x_{1}+\rmi y_{1}$. Then \Eq{A2.eq} becomes the
condition that the four real variables describing a qubit denote a
point on the three-sphere $\Sphere{3}$ (a 3-manifold) embedded in
$\R^{4}$: 
\begin{equation}
{x_{0}}^{2}+{y_{0}}^{2}+{x_{1}}^{2}+{y_{1}}^{2}=1\,.\label{A5.eq}
\end{equation}

We can reduce 3 real degrees of freedom in \Eq{A5.eq} to 2 real
degrees of freedom by effectively removing $\rme^{\rmi\theta}$ (``fibering
out by the circle $\Sphere{1}$''). The standard form of this maps
(``the Hopf fibration'') is 
\begin{eqnarray}
X & = & 2\,\mathrm{Re}\ \alpha_{0}\alpha_{1}^{*}=2x_{0}x_{1}+2y_{0}y_{1}\,,\nonumber \\
Y & = & 2\,\mathrm{Im}\ \alpha_{0}\alpha_{1}^{*}=2x_{1}y_{0}-2x_{0}y_{1}\,,\label{eq:HopfFibration}\\
Z & = & \left|\alpha_{0}\right|^{2}-\left|\alpha_{1}\right|^{2}={x_{0}}^{2}+{y_{0}}^{2}-{x_{1}}^{2}-{y_{1}}^{2}\,.\nonumber 
\end{eqnarray}
By denoting the the three-dimensional vector~$\left(X,Y,Z\right)$
as $\Hat{a}$, \Eq{A5.eq} implies these transformed coordinates
obeying 
\begin{equation}
\left\Vert \Hat{a}\right\Vert ^{2}=X^{2}+Y^{2}+Z^{2}=\left(\left|\alpha_{0}\right|^{2}+\left|\alpha_{1}\right|^{2}\right)^{2}=1\label{A7.eq}
\end{equation}
and therefore have only two remaining degrees of freedom describing
all possible distinct one-qubit quantum states. In Fig.~\ref{hopfFibers.fig},
we illustrate schematically the family of circles \textit{each one
of which is collapsed to a point}~$\left(\phi,\psi\right)$ on the
surface $X^{2}+Y^{2}+Z^{2}=1$ by the Hopf map.

\begin{figure}[tb]
\hfill{}\subfloat[]{\centering{}\includegraphics[width=0.48\columnwidth]{hopfS2}}\hfill{}\subfloat[]{\begin{centering}
\includegraphics[width=0.48\columnwidth]{hopfFibers}
\par\end{centering}
}\hfill{} \caption{(a) The sphere represented by \Eq{A7.eq}, which is the irreducible
space of one-qubit states, along with a representative set of points
on the sphere. (b) Representation of the Hopf fibration, Eq.~(\ref{eq:HopfFibration}),
as a family of circles (the paths of $\rme^{\rmi\theta}$) on the
three-sphere $\Sphere{3}$. Although $\Sphere{3}$ cannot be directly
embedded in $\R^{3}$, $\Sphere{3}$ can be regarded as attaching
two three-dimensional ball on two sides of two-sphere $\Sphere{2}$.
In this way, each circle in $\Sphere{3}$ corresponding to a single
point on the sphere in (a) can be represented as a circle in the three-dimensional
ball. Moreover, points in (a) are color coded corresponding to circles
in (b), e.g., one pole contains the red elliptical circle that would
become an infinite-radius circle by a slightly different way to represent
$\Sphere{3}$ in $\R^{3}$, and the opposite pole corresponds to the
large perfectly round red circle at the equator.}
\label{hopfFibers.fig} 
\end{figure}

The resulting manifold is the two-sphere $\Sphere{2}$ (a 2-manifold)
embedded in $\R^{3}$. If we choose one of many possible coordinate
systems describing $\Sphere{3}$ via \Eq{A5.eq} such as 
\begin{equation}
\left(x_{0},y_{0},x_{1},y_{1}\right)=\left(\cos\left(\theta+\phi\right)\cos\psi,\,\sin\left(\theta+\phi\right)\cos\psi,\,\cos\left(\theta-\phi\right)\sin\psi,\,\sin\left(\theta-\phi\right)\sin\psi\right)\,,\label{A8a.eq}
\end{equation}
where %% $0\leq \theta \leq 2 \pi$,  $-\pi \leq \phi \leq  \pi$, 
$0\leq\psi\leq\frac{\pi}{2}$, with $0\leq\theta+\phi<2\pi$ and $0\leq\theta-\phi<2\pi$,
we see that 
\begin{equation}
\left(X,Y,Z\right)=\left(\cos\left(2\phi\right)\sin\left(2\psi\right),\,\sin\left(2\phi\right)\sin\left(2\psi\right),\,\cos\left(2\psi\right)\right)\,.\label{A8b.eq}
\end{equation}
Thus the one-qubit state is independent of $\theta$, and we can choose
$\theta=\phi$ without loss of generality, reducing the form of the
unique one-qubit states to $\ket{\psi_{1}}=\rme^{2\rmi\phi}\cos\psi\ket{0}+\sin\psi\ket{1}$,
and an irreducible state can be represented as a point on a sphere
called the Bloch sphere, as shown in Fig.~\ref{blochSphere.fig}(a).

\begin{figure}[tb]
\begin{centering}
\hfill{}\subfloat[]{\centering{}\includegraphics[width=0.47\columnwidth]{blochSphere}}\hfill{}\subfloat[]{\begin{centering}
\includegraphics[width=0.47\columnwidth]{blochSphereArc}
\par\end{centering}
}\hfill{}
\par\end{centering}
\caption{(a) The conventional Bloch sphere with a unique state represented
by the point at the red sphere. (b) The geodesic shortest-distance
arc connecting two one-qubit quantum states.}
\label{blochSphere.fig} 
\end{figure}

@@

For one qubit, we could easily solve the problem of reducing the full
unit-norm space to its irreducible components $\Vec{X}=(X,Y,Z)$ characterizing
the Bloch sphere. We have just argued that essentially the same process
is possible for $n$-qubits: in the abstract argument, we simply identify
the family of coefficients $\{\alpha_{i}\}$ as being the same if
they differ only by an overall phase $\rme^{\rmi\theta}$. However,
in practice this is not a construction that is easy to realize in
a practical computation. We now outline an explicit algorithm for
accomplishing the reduction to the irreducible $n$-qubit state space
$\CP{D-1}$; this construction will turn out to be useful for the
validation of our discrete results to follow below.

@@

We can now use the complex generalization of the classical Veronese
coordinate system for projective geometry to remove the overall phase
ambiguity $\rme^{\rmi\theta}$ from the $n$-qubit states. If we take
a particular weighting of the elements of the density matrix $\rho$,
we can construct a \textit{unit vector\/} of real dimension $D^{2}$
with the form: 
\begin{equation}
\Vec{X}=(|\alpha_{i}|^{2},\ldots,\sqrt{2}\,\mathrm{Re}\ \alpha_{i}\alpha_{j}^{*},\ldots,\sqrt{2}\,\mathrm{Im}\ \alpha_{i}\alpha_{j}^{*},\ldots)\ ,\label{nqCoords.eq}
\end{equation}
where 
\begin{eqnarray}
\Vec{X}\cdot\Vec{X}=\left(\sum_{i=0}^{D-1}\left|\alpha_{i}\right|^{2}\right)^{2}=1\ .
\end{eqnarray}
This construction gives an explicit embedding of the $(D-1)$-dimensional
complex, or $(2D-2)=(2^{n+1}-2)$-dimensional real, object in a real
space of dimension $D^{2}=2^{2n}$. However, this is somewhat subtle
because the vector is of unit length, so technically the embedding
space is a sphere of dimension $D^{2}-1=2^{2n}-1$ embedded in $\R^{D^{2}}$;
the one-qubit irreducible states could be represented in a 4D embedding,
but the magnitude of every coordinate would be one; furthermore, the
object embedded in the resulting $\Sphere{3}$ is indeed $\Sphere{2}$
because we can fix one complex coordinate to be unity, and let one
vary, giving a total of two irreducible dimensions. In fact one must
choose \textit{two\/} coordinate patches, one covering one pole of
$\Sphere{2}$ with coordinates 
\begin{eqnarray}
\begin{array}{ccc}
\alpha_{0} & = & 1+0i\\
\alpha_{1} & = & x_{1}+iy_{1}
\end{array}
\end{eqnarray}
and the other patch covering the other pole of $\Sphere{2}$ with
coordinates 
\begin{eqnarray}
\begin{array}{ccc}
\alpha_{0} & = & x_{0}+iy_{0}\\
\alpha_{1} & = & 1+0i\ .
\end{array}
\end{eqnarray}
